---
title: "STATS 326/786"
author: "Dates, Periods, Durations and Intervals with lubridate"
date: \today
fontsize: 11pt
output:
  bookdown::html_document2:
    fig_height: 5
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(lubridate)
library(tidyverse)
```

# Overview

This walkthrough introduces the `lubridate` package for working with dates and times in R.  We cover four core topics:

1. **Date and datetime creation** — parsing strings, constructing from components, and time zones.
2. **Periods** — human-friendly time spans (e.g., "1 month", "3 years").
3. **Durations** — exact physical time spans in seconds.
4. **Intervals** — a specific span anchored between two instants.

---

# 1 Date and Datetime Creation

## 1.1 Parsing strings

`lubridate` provides a family of `ymd()`, `dmy()`, `mdy()` (and datetime variants) that parse strings flexibly.

```{r parse-strings}
# Year-month-day
ymd("2024-03-14")
ymd("20240314")       # No separators also works

# Day-month-year
dmy("14/03/2024")
dmy("14 March 2024")

# Month-day-year (US format)
mdy("March 14, 2024")
mdy("03-14-2024")
```

## 1.2 Adding a time component

Append `_hms`, `_hm`, or `_h` to include a time.

```{r parse-datetimes}
# UTC by default
dt_utc <- ymd_hms("2024-03-14 09:30:00")
dt_utc

# Specify a time zone
dt_nz <- ymd_hms("2024-03-14 09:30:00", tz = "Pacific/Auckland")
dt_nz

# Converting between time zones (same instant, different representation)
with_tz(dt_nz, tzone = "UTC")
```

## 1.3 Constructing from components

`make_date()` and `make_datetime()` build a date/datetime from individual numeric parts.

```{r make-date}
# Build a date
make_date(year = 2024, month = 3, day = 14)

# Build a datetime with time zone
make_datetime(year  = 2024,
              month = 3,
              day   = 14,
              hour  = 9,
              min   = 30,
              sec   = 0,
              tz    = "Pacific/Auckland")
```

## 1.4 Extracting components

Once you have a date or datetime object, accessor functions let you pull out individual parts.

```{r accessors}
d <- ymd_hms("2024-07-04 18:45:30", tz = "US/Eastern")

year(d)
month(d)
month(d, label = TRUE)   # Abbreviated month name
day(d)
wday(d, label = TRUE)    # Day of the week
hour(d)
minute(d)
second(d)
tz(d)                    # Time zone
```

You can also *set* components by assigning into these functions:

```{r set-components}
d2 <- d
year(d2) <- 2025
d2
```

## 1.5 Today and now

```{r today-now}
today()          # Current date (no time)
now()            # Current datetime (system time zone)
now("UTC")       # Current datetime in UTC
```

---

# 2 Periods

A **period** is a human-calendar time span.  It is specified in "clock" units — years, months, days, hours, minutes, seconds — as a person would describe them.  Adding a period of `months(1)` to 31 January gives 28 February (the end of the next month), respecting calendar irregularities like leap years and varying month lengths.

## 2.1 Creating periods

```{r create-periods}
years(1)
months(6)
weeks(2)
days(10)
hours(3)
minutes(30)
seconds(45)

# Combine with +
years(1) + months(6) + days(10)
```

## 2.2 Arithmetic with periods

```{r period-arithmetic}
start <- ymd("2024-01-31")

# Adding one month: period respects calendar
start + months(1)   # 2024-02-29 (2024 is a leap year)

# Compare with non-leap year
ymd("2023-01-31") + months(1)   # 2023-02-28

# Adding years across a leap year boundary
ymd("2024-02-29") + years(1)    # 2025-02-29 does not exist -> NA
ymd("2024-02-29") + years(4)    # 2028-02-29 exists -> valid

# Arithmetic on datetimes
dt <- ymd_hms("2024-03-10 02:00:00", tz = "US/Eastern")
dt + hours(3)
dt + days(1) + hours(3)
```

## 2.3 Periods and DST

Because periods follow the clock, they can cross Daylight Saving Time (DST) transitions naturally.

```{r periods-dst}
# US/Eastern clocks spring forward 2024-03-10 02:00 -> 03:00
before_dst <- ymd_hms("2024-03-10 01:00:00", tz = "US/Eastern")
before_dst + hours(2)   # Period: 03:00 (clock jumps, only 1 real hour passes)
```

---

# 3 Durations

A **duration** is an exact physical time span measured in seconds.  Unlike periods, durations do not account for calendar irregularities.  Adding `dyears(1)` always adds exactly 31,557,600 seconds (365.25 × 24 × 60 × 60), regardless of whether it crosses a leap year or a DST change.

## 3.1 Creating durations

Duration constructors are prefixed with `d`:

```{r create-durations}
dyears(1)        # 31,557,600 seconds
dmonths(1)       # 2,629,800 seconds (1/12 of a year)
dweeks(1)        # 604,800 seconds
ddays(1)         # 86,400 seconds
dhours(1)        # 3,600 seconds
dminutes(1)      # 60 seconds
dseconds(1)      # 1 second

# Convert a difftime directly to a duration
as.duration(difftime(ymd("2024-12-31"), ymd("2024-01-01")))
```

## 3.2 Arithmetic with durations

```{r duration-arithmetic}
start <- ymd("2024-01-31")

# Exact seconds added — result may differ from period arithmetic
start + dmonths(1)   # Not necessarily end of February

# Duration in years from Einstein's lifespan
birth <- make_date(1879, 3, 14)
death <- make_date(1955, 4, 18)

(death - birth)            # difftime in days
(death - birth) / ddays(1) # Number of days as a plain number
(death - birth) / dyears(1) # Convert to years
```

## 3.3 Periods vs Durations — side-by-side comparison

```{r compare}
ref <- ymd_hms("2024-03-09 12:00:00", tz = "US/Eastern")  # Day before DST

# Period: follows the clock
ref + days(1)

# Duration: adds exactly 86,400 seconds (loses an hour due to DST spring-forward)
ref + ddays(1)
```

> **Rule of thumb:**  Use *periods* when you want human-readable calendar arithmetic (e.g., "add 1 month").  Use *durations* when you need exact elapsed time in seconds (e.g., computing flight transit time).

---

# 4 Intervals

An **interval** is a time span anchored between a specific *start* and *end* instant.  Unlike periods or durations, an interval carries its position in time, which allows exact calendar- and DST-aware calculations.

## 4.1 Creating intervals

```{r create-intervals}
start <- ymd_hms("2024-03-14 09:00:00", tz = "Pacific/Auckland")
end   <- ymd_hms("2024-03-16 17:30:00", tz = "Pacific/Auckland")

int <- interval(start, end)
int

# Shorthand using %--% operator
int2 <- start %--% end
int2

# Check they are the same
identical(int, int2)
```

## 4.2 Interval length

Convert an interval to a duration or period to get its length:

```{r interval-length}
# As a duration (in seconds by default)
as.duration(int)

# In convenient units
int / dminutes(1)   # Minutes
int / dhours(1)     # Hours
int / ddays(1)      # Days

# As a period (calendar-aware)
as.period(int)
```

## 4.3 Time-zone-aware intervals

```{r tz-interval}
# Flight: Auckland to Paris
depart <- make_datetime(2022, 11, 18, 15, 15, tz = "Pacific/Auckland")
arrive <- make_datetime(2022, 11, 19, 12, 20, tz = "Europe/Paris")

flight_int <- interval(depart, arrive)
flight_int

# Duration in minutes
flight_int / dminutes(1)

# Duration in hours
flight_int / dhours(1)
```

## 4.4 Testing membership with `%within%`

```{r within}
event <- ymd_hms("2024-03-15 12:00:00", tz = "Pacific/Auckland")

event %within% int   # Is this instant inside the interval?

# A vector of dates
dates <- ymd(c("2024-03-13", "2024-03-14", "2024-03-15", "2024-03-17"))
dates %within% int
```

## 4.5 Overlapping and adjacent intervals

```{r overlap}
int_a <- interval(ymd("2024-01-01"), ymd("2024-06-30"))
int_b <- interval(ymd("2024-04-01"), ymd("2024-12-31"))
int_c <- interval(ymd("2024-07-01"), ymd("2024-12-31"))

int_overlaps(int_a, int_b)   # TRUE — they share April–June
int_overlaps(int_a, int_c)   # FALSE — they share no days (touch at boundary)
```

---

# 5 Summary Table

| Concept | Constructor example | Key property |
|---|---|---|
| Date | `ymd("2024-03-14")` | Calendar date, no time |
| Datetime | `ymd_hms("2024-03-14 09:00:00", tz = "UTC")` | Instant in time |
| Period | `months(1) + days(3)` | Calendar units, follows clock |
| Duration | `dmonths(1) + ddays(3)` | Exact seconds, physical time |
| Interval | `interval(start, end)` | Anchored span between two instants |

---

# 6 Quick Reference

```{r quick-reference, eval=FALSE}
# Parsing
ymd(); dmy(); mdy()             # Date only
ymd_hms(); dmy_hm()             # Date + time
make_date(); make_datetime()     # From numeric components

# Accessors
year(); month(); day()
hour(); minute(); second()
wday(); yday(); week()
tz(); with_tz(); force_tz()

# Periods (calendar)
years(); months(); weeks(); days(); hours(); minutes(); seconds()

# Durations (exact seconds)
dyears(); dmonths(); dweeks(); ddays(); dhours(); dminutes(); dseconds()

# Intervals
interval(start, end)  or  start %--% end
int_overlaps(); %within%
as.duration(); as.period()
```
