---
title: "STATS 326/786 — Lab 02 Q3(a) Model Answer"
author: "Week 2 Tutorial"
date: \today
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(lubridate)
library(tidyverse)
library(nycflights13)
```

# Problem 3 — Setup

Filter the `flights` dataset to all JFK → LAX flights with a recorded arrival time.

```{r}
data <- flights %>%
  filter(origin == "JFK",
         dest   == "LAX",
         !is.na(arr_time))
```

# Problem 3(a)

> Create a departure date-time variable by considering the scheduled departure time and departure delay.  Note that `time_hour` is the date-time variable for the scheduled departure time.

## Solution

```{r}
data <- data %>%
  mutate(dep_datetime = time_hour + minutes(dep_delay))
```

## Explanation

| Component | Role |
|---|---|
| `time_hour` | A `POSIXct` date-time giving the **scheduled** departure hour (e.g., `2013-01-01 05:00:00 EST`) |
| `dep_delay` | Departure delay in **minutes** (negative = early departure) |
| `minutes(dep_delay)` | A `lubridate` period object converting the integer delay into minutes |
| `time_hour + minutes(dep_delay)` | Adds the delay to the scheduled time to give the **actual** departure date-time |

**Why `minutes()` and not just `+ dep_delay`?**

Adding a plain integer to a `POSIXct` object adds **seconds**, not minutes.
`lubridate::minutes()` creates a period of the correct unit, making the arithmetic unambiguous and self-documenting.

**Note on time zones:** `time_hour` in `nycflights13` is already in `US/Eastern`
(JFK's local time zone), so no explicit `tz` conversion is needed for the
departure datetime.

## Quick check

```{r}
data %>%
  select(time_hour, dep_delay, dep_datetime) %>%
  head(5)
```
