---
title: "STATS 326/786 — Lab 02 Q3(b) Model Answer"
author: "Week 2 Tutorial"
date: \today
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(lubridate)
library(tidyverse)
library(nycflights13)
```

# Problem 3 — Setup

Filter the `flights` dataset to all JFK → LAX flights with a recorded arrival time,
and recreate the departure date-time from Q3(a).

```{r}
data <- flights %>%
  filter(origin == "JFK",
         dest   == "LAX",
         !is.na(arr_time))

# Q3(a): actual departure date-time
data <- data %>%
  mutate(dep_datetime = time_hour + minutes(dep_delay))
```

# Problem 3(b)

> Create an arrival date-time variable by considering the actual arrival time,
> the date of departure, and whether a flight arrives on the same day or the
> following day (overnight flights).
> Hint: Convert `arr_time` into a character variable of length 4 using `sprintf`.
> Then extract the hour and minutes using `substr` when creating your arrival
> date-time variable.  Then create an indicator to determine which flights are
> overnight (1) or not (0) and add this to your arrival date-time.

## Solution

### Step 1 — Parse `arr_time` into hours and minutes

`arr_time` is stored as an integer (e.g., `836` means 08:36, `2358` means 23:58).
We need to split it into separate hour and minute components.

```{r}
data <- data %>%
  mutate(
    arr_time_char = sprintf("%04d", arr_time),               # zero-pad to 4 digits
    arr_datetime  = make_datetime(
      year  = year,
      month = month,
      day   = day,
      hour  = as.numeric(substr(arr_time_char, 1, 2)),       # first 2 chars = hour
      min   = as.numeric(substr(arr_time_char, 3, 4)),       # last  2 chars = minute
      tz    = "US/Pacific"                                   # LAX is US/Pacific
    )
  )
```

**Key steps explained:**

| Code | Purpose |
|---|---|
| `sprintf("%04d", arr_time)` | Converts integer to a 4-character string, left-padding with zeros (e.g., `56` → `"0056"`) |
| `substr(arr_time_char, 1, 2)` | Extracts characters 1–2 → the **hour** |
| `substr(arr_time_char, 3, 4)` | Extracts characters 3–4 → the **minute** |
| `make_datetime(..., tz = "US/Pacific")` | Assembles the date-time using the **departure date** (`year`, `month`, `day`) and the LAX time zone |

At this point `arr_datetime` assumes every flight arrives on the **same calendar
date** as its departure.  That is not always true.

### Step 2 — Adjust for overnight flights

A flight is overnight if it arrives (in Pacific time) before it departs (in
Eastern time) when both are compared on a common clock.  We detect this by
checking whether `arr_datetime < dep_datetime` and add one day where needed.

```{r}
data <- data %>%
  mutate(overnight    = arr_datetime < dep_datetime) %>%     # TRUE/FALSE → coerces to 1/0
  mutate(arr_datetime = arr_datetime + days(overnight))      # add 0 or 1 day
```

`days(overnight)` works because `TRUE` coerces to `1` and `FALSE` to `0`, so
`days(1)` is added for overnight flights and `days(0)` (no change) for same-day
arrivals.

## Quick check

```{r}
data %>%
  select(dep_datetime, arr_datetime, overnight, air_time) %>%
  head(10)
```

Verify that `arr_datetime` is always later than `dep_datetime` after the
adjustment:

```{r}
any(data$arr_datetime < data$dep_datetime)   # Should be FALSE
```
