---
title: "STATS 326/786 — Lab 02 Q3(c) Model Answer"
author: "Week 2 Tutorial"
date: \today
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
library(lubridate)
library(tidyverse)
library(nycflights13)
```

# Problem 3 — Setup

Recreate the departure and arrival date-times from Q3(a) and Q3(b).

```{r}
data <- flights %>%
  filter(origin == "JFK",
         dest   == "LAX",
         !is.na(arr_time))

# Q3(a): actual departure date-time
data <- data %>%
  mutate(dep_datetime = time_hour + minutes(dep_delay))

# Q3(b): actual arrival date-time (US/Pacific), adjusted for overnight flights
data <- data %>%
  mutate(
    arr_time_char = sprintf("%04d", arr_time),
    arr_datetime  = make_datetime(
      year  = year,
      month = month,
      day   = day,
      hour  = as.numeric(substr(arr_time_char, 1, 2)),
      min   = as.numeric(substr(arr_time_char, 3, 4)),
      tz    = "US/Pacific"
    )
  ) %>%
  mutate(overnight    = arr_datetime < dep_datetime) %>%
  mutate(arr_datetime = arr_datetime + days(overnight))
```

# Problem 3(c)

> We are told that the average transit time is 6.5 hours.  Create a variable to
> calculate the transit time, i.e., the difference between the arrival date-time
> and departure date-time.  Plot the distribution and comment.  Then plot the
> transit time against the air time and comment.

## Solution

### Calculate transit time

```{r}
data <- data %>%
  mutate(transit_time = arr_datetime - dep_datetime)
```

`arr_datetime - dep_datetime` returns a `difftime` object in hours (since both
are `POSIXct` with different time zones, lubridate handles the UTC conversion
automatically).

### Plot 1 — Distribution of transit time

```{r, message = FALSE, warning = FALSE}
ggplot(data = data,
       mapping = aes(x = transit_time)) +
  geom_histogram(fill = "blue", col = "white") +
  geom_vline(xintercept = 6.5, linewidth = 1, linetype = "dashed") +
  theme_bw() +
  labs(x = "Transit time (hours)",
       y = "Count")
```

**Comment:** The distribution of transit times is roughly unimodal and
approximately symmetric, centred close to the claimed average of 6.5 hours
(dashed line).  The bulk of flights have transit times between about 5.5 and
7.5 hours.  There is a small number of outliers with unusually long transit
times, likely due to large arrival delays.

### Plot 2 — Transit time vs air time

```{r, message = FALSE, warning = FALSE}
ggplot(data = data,
       mapping = aes(x = transit_time,
                     y = air_time / 60)) +
  geom_point(alpha = 0.1, col = "blue", size = 0.5) +
  geom_abline(slope = 1, intercept = 0) +
  theme_bw() +
  labs(x = "Transit time (hours)",
       y = "Air time (hours)")
```

**Comment:** Transit time is always **greater** than air time — all points lie
below the 45° reference line (slope = 1, intercept = 0).  This is expected:
`arr_time` in the dataset records when the aircraft arrives at the gate, not
when it touches down, so transit time includes taxi time at both ends plus any
time spent holding before landing.  The two variables are strongly positively
correlated, with the gap between them representing the ground/taxi component of
the journey.
