---
title: "STATS 326/786"
author: "Week 2 Tutorial"
date: \today
fontsize: 11pt
output:
  bookdown::html_document2:
    fig_height: 5
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(lubridate)
library(tidyverse)
```

# Problem 1

> Albert Einstein was born on 14 March 1879 and died on 18 April 1955.  Use `lubridate` functions to calculate how many days he was alive for.  Then calculate this as a "duration" in years.  You may ignore time-zones for this question.

```{r}

birth <- make_date(year = 1879, month = 3, day = 14)
death <- make_date(year = 1955, month = 4, day = 18)

(death - birth)/dyears(1)                   
                   
```


# Problem 2

> Matt went to a conference in Toulouse in 2022.  He departed Auckland at 1515 on 18 November 2022, and landed in Toulouse, France at 1220 on 19 November.  Use `lubridate` functions to create a time interval object from departure to landing.  Then compute the duration (in minutes) he was in transit for.  Be careful with time-zones.

```{r}

dep_datetime <- make_datetime(year = 2022, 
                              month = 11, 
                              day = 18, 
                              hour = 15, 
                              min = 15, 
                              tz = "Pacific/Auckland")

arr_datetime <- make_datetime(year = 2022,
                              month = 11,
                              day = 19,
                              hour = 12,
                              min = 20,
                              tz = "Europe/Paris")

interval <- lubridate::interval(start = dep_datetime, end = arr_datetime)

interval/dminutes(60)

```

# Problem 3

> The `flights` data set in the `nycflights13` package contains information on all flights from New York in 2013.  Install and load this `R` package and type `?flights` to look at the help file to see variable details.  Consider all flights from JFK (US/Eastern time-zone) to LAX (US/Pacific time-zone).  

```{r}

install.packages("nycflights13")
library(nycflights13)

head(flights)

```

## (a) 

> Create a departure date-time variable by considering the scheduled departure time and departure delay.  Note that `time_hour` is the date-time variable for the scheduled departure time.  

```{r}

data <- flights %>%
  filter(origin == "JFK",
         dest   == "LAX",
         !is.na(arr_time))


data <- data %>% 
  mutate(actual_dep_time = time_hour + minutes(dep_delay))

glimpse(data)


```


## (b)

> Create an arrival date-time variable by considering the actual arrival time, the date of departure, and whether a flight arrives on the same day or the following day (overnight flights). 
Hint: Convert `arr_time` into a character variable of length 4 using `sprintf`.  Then extract the hour and minutes using `substr` when creating your arrival date-time variable.  Then create an indicator to determine which flights are overnight (1) or not (0) and add this to your arrival date-time.  

```{r}

```


## (c)

> We are told that the average transit time is 6.5 hours.  Create a variable to calculate the transit time, i.e., the difference between the arrival date-time and departure date-time.  Plot the distribution and comment. Then plot the transit time against the air time and comment.  

```{r, message = FALSE, warning = FALSE}

```


