---
title: "STATS 326/786"
author: "Week 4 Tutorial"
date: \today
fontsize: 11pt
output:
  bookdown::html_document2:
    fig_height: 5
    toc: yes
    toc_depth: 1
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(fpp3)
```


# Problem 1

> Below is some fake seasonal data with a perfectly periodic pattern, no trend-cycle, and no remainder components.  In class, we learned that moving averages (MA) are useful for smoothing time series and estimating a trend-cycle. A $2 \times 4~\text{MA}$ and a $5~\text{MA}$ have been overlaid on the original time series.  What do you notice about the MAs, and why might this be happening?  

```{r, warning = FALSE, echo = FALSE}

data <- tsibble(Quarter = yearquarter(1:40),
                Data = rep(c(4, 9, 3, 7), 10),
                index = Quarter)

data.ma <- data %>%
  mutate(`4-MA` = slider::slide_dbl(Data, 
                                    mean, 
                                    .before = 1, 
                                    .after = 2, 
                                    .complete = TRUE),
         `2x4-MA` = slider::slide_dbl(`4-MA`, 
                                      mean, 
                                      .before = 1, 
                                      .after = 0, 
                                      .complete = TRUE),
         `5-MA` = slider::slide_dbl(Data, 
                                    mean, 
                                    .before = 2,
                                    .after = 2, 
                                    .complete = TRUE))

data.long = data.ma %>%
  pivot_longer(cols = c('Data', "2x4-MA", "5-MA"),
               names_to = "Series", values_to = "Measurement")

ggplot(data = data.long) + 
  geom_line(mapping = aes(x = Quarter, y = Measurement, col = Series)) + 
  theme_bw()
  
```

> Repeat the above exercise, but for the fake data below.  This data is seasonal (again, with a perfectly periodic pattern) with a linear trend and no remainder component.  

```{r, warning=FALSE, echo = FALSE}

data <- tsibble(Quarter = yearquarter(1:40),
                y = rep(c(4, 9, 3, 7), 10) + 0.5 * 1:40,
                index = Quarter)


data %>% 
  autoplot(y, colour = "grey") + 
  theme_minimal()

```

```{r}

data_ma <- data |> 
  mutate(`4-MA`= slider::slide_dbl(y,
                                   mean,
                                   .before = 1,
                                   .after = 2,
                                   .complete = TRUE),
         `2x4-MA`= slider::slide_dbl(`4-MA`,
                                      mean,
                                      .before = 1,
                                      .after = 0,
                                      .complete = TRUE),
         `5-MA`= slider::slide_dbl(y, 
                                    mean,
                                    .before = 2,
                                    .after = 2,
                                    .complete = TRUE)
         )
  
data_ma |> 
  autoplot(y, colour = "grey") +
  autolayer(data_ma, `2x4-MA`, col = 2) +
  autolayer(data_ma, `5-MA`, col = 4)+
  theme_minimal()


```



> If you were to apply a $3 \times 5~\text{MA}$ to a time series, analytically derive the weights that each observation contributes to the moving-averaged time series at time $t$. 





# Problem 2

> Below is an example of a time series with multiplicative seasonality.  In this problem, you learn how to perform your own (multiplicative) classical decomposition from first principles.  

```{r, echo = FALSE}
a10 <- PBS %>%
  filter(ATC2 == "A10") %>%
  select(Month, Concession, Type, Cost) %>%
  summarise(Total_Cost = sum(Cost)) %>%
  mutate(Total_Cost = Total_Cost / 1e6,
         Month.Name = month(Month, label = TRUE)) 

a10 %>%
  autoplot(Total_Cost) + 
  theme_minimal()
```

> Calculate the trend-cycle by performing a $2\times 12~\text{MA}$ on your time series.

```{r}

a10 <- a10 |> 
  mutate(`12-MA`= slider::slide_dbl(Total_Cost,
                                    mean,
                                    .before = 5,
                                    .after = 6,
                                    .complete = TRUE),
         Trend = slider::slide_dbl(`12-MA`,
                                      mean,
                                      .before = 1,
                                      .after = 0,
                                      .complete = TRUE)
         )


a10 |> autoplot(Total_Cost) +
  autolayer(a10, Trend, col = 4)


```
         
> Compute the detrended time series (original time series divided by trend).    
         
```{r}         

a10 <- a10 |> 
  mutate(Detrended = Total_Cost / Trend)

a10 |> autoplot(Total_Cost) +
  autolayer(a10, Trend, col = 2) +
  autolayer(a10, Detrended, col = 4)


glimpse(a10)


```

> Calculate the seasonal factor for each month (noting that these must sum to 12 for a monthly time series).  

```{r}

a10_s <- a10 |> 
  index_by(Month.Name) |> 
  summarise(Seasonal = mean(Detrended, na.rm = TRUE)) |> 
  glimpse()

seasonal <- a10_s |> pull(Seasonal) |> print()

seasonal |> prod()

sum_s

# small adjustment to seasonal factors so that they sum to 12
# 
a10_s <- a10_s |> 
  mutate(Seasonal = Seasonal / sum_s * 12) |> 
  print()

prod(a10_s$Seasonal)

glimpse(a10_s)

```

> Merge your data sets so the seasonal factor is now included in your original tsibble.

```{r}

```

> Compute the remainder component (as detrended divided by seasonal) and seasonally-adjusted component (as original time series divided by seasonal).

```{r}

```

> Plot the original time series and overlay it with the trend-cycle and seasonally-adjusted series.  

```{r, warning = FALSE}

```

> Extra: See if you can re-create the classical decomposition plot.

```{r, warning = FALSE}

```
